---
title: "Portfolio Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{portfolio-functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits=3)
```

This vignette illustrates the portfolio analysis functions in the **introcompfinr** package. The vignette is based on Chapter 11 of *Introduction to Computational Finance and Financial Econometrics with R* by Eric Zivot.

```{r setup}
library(introcompfinr)
```

## Portfolio Analysis Functions in R 
The package **introcompfinr** contains a few R functions for
computing Markowitz mean-variance efficient portfolios allowing for
short sales using matrix algebra computations. These functions allow
for the easy computation of the global minimum variance portfolio,
an efficient portfolio with a given target expected return, the tangency
portfolio, and the efficient frontier. These functions are summarized
in Table below.

------------------------------------------------------------------------------------------
**Function**                **Description**
------------------------    --------------------------------------------------------------
`get_portfolio`             create portfolio object 
  
`globalmin_portfolio`       compute global minimum variance portfolio 
  
`efficient_portfolio`       compute minimum variance portfolio subject to target return 
  
`tangency_portfolio`        compute tangency portfolio 
  
`efficient_frontier`        compute efficient frontier of risky assets 
------------------------------------------------------------------------------------------




The following examples illustrate the use of the functions in listed in the table
with example data from Chapter 11 of *Introduction to Computational Finance and Financial Econometrics with R*. 

```{r echo=TRUE}
asset.names <- c("MSFT", "NORD", "SBUX")
mu.vec = c(0.0427, 0.0015, 0.0285)
sigma.mat = matrix(c(0.0100, 0.0018, 0.0011,
           0.0018, 0.0109, 0.0026,
           0.0011, 0.0026, 0.0199),
         nrow=3, ncol=3)
r.f=0.005
```

To specify a `portfolio` object, you need an expected
return vector $\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$ for the assets under consideration
as well as a vector of portfolio weights $\mathbf{x}$. To create an equally weighted
portfolio use:

```{r echo=TRUE}

ew = rep(1,3)/3
equalWeight.portfolio = get_portfolio(er_vec=mu.vec,cov_mat=sigma.mat,weight_vec=ew)
class(equalWeight.portfolio)
```

Portfolio objects have the following components:

```{r echo=TRUE}

names(equalWeight.portfolio)
```

There are `print()`, `summary()` and `plot()`
methods for portfolio objects. The `print()` method gives:

```{r echo=TRUE}

equalWeight.portfolio
```

The `plot()` method shows a bar chart of the portfolio weights:

```{r PortMatrixPlotMethod, fig.cap='Plot method for objects of class portfolio', out.width='70%', fig.width = 6, fig.asp=.618, fig.align='center', echo=TRUE}

plot(equalWeight.portfolio, col="cornflowerblue")
```


The resulting plot is shown in Figure \@ref(fig:PortMatrixPlotMethod).
The global minimum variance portfolio (allowing for short sales) $\mathbf{m}$
solves the optimization problem 

$$
\min_{\mathbf{m}}~\sigma_{p,m}^{2}=\mathbf{m^{\prime}{\Sigma} m} ~ s.t. ~ \mathbf{m^{\prime}1}=1.
$$

and is computed using 

$$
\mathbf{m}= \frac{\boldsymbol{\Sigma}^{-1}\mathbf{1}}{\mathbf{1}^{\prime}{\boldsymbol{\Sigma}}^{-1}\mathbf{1}}.
$$
To compute this portfolio use the function `globalmin_portfolio()`:

```{r echo=TRUE}

gmin.port = globalmin_portfolio(mu.vec, sigma.mat)
class(gmin.port)
gmin.port
```

A mean-variance efficient portfolio $\mathbf{x}$ that achieves the
target expected return $\mu_{0}$ solves the optimization problem

$$
\begin{aligned}
\min_{\mathbf{x}}~\sigma_{p,x}^{2} & =\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x} \\
\textrm{ s.t.  } \mu_{p} & =\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p,0}\mathbf{,}\textrm{ and }\mathbf{x}^{\prime}\mathbf{1}=1. 
\end{aligned}
$$

and is computed by solving the system of linear equations $\mathbf{Az}_{x}\mathbf{=b}_{0}$, where 
$$
\mathbf{A}=\left(\begin{array}{ccc}
2\boldsymbol{\Sigma} & \boldsymbol{\mu} & \mathbf{1}\\
\boldsymbol{\mu}^{\prime} & 0 & 0\\
\mathbf{1}^{\prime} & 0 & 0
\end{array}\right),~\mathbf{z}_{x}=\left(\begin{array}{c}
\mathbf{x}\\
\lambda_{1}\\
\lambda_{2}
\end{array}\right)\textrm{ and }\mathbf{b}_{0}=\left(\begin{array}{c}
\mathbf{0}\\
\mu_{p,0}\\
1
\end{array}\right).
$$
The solution for $\mathbf{z}_{x}$ is: 
$$
\mathbf{z}_{x}=\mathbf{A}^{-1}\mathbf{b}_{0}.
$$
The first $N$ elements of $\mathbf{z}_{x}$ are the portfolio weights
$\mathbf{x}$ for the minimum variance portfolio with expected return
$\mu_{p,x}=\mu_{p,0}$. If $\mu_{p,0}$ is greater than or equal to
the expected return on the global minimum variance portfolio, then
$\mathbf{x}$ is an efficient (frontier) portfolio. Otherwise, it
is an inefficient (frontier) portfolio. 


To compute this portfolio for the target expected return $\mu_{0}=E[R_{\textrm{msft}}]=0.0427$
use the `efficient_portfolio()` function:

```{r echo=TRUE}

target.return = mu.vec[1]
e.port.msft = efficient_portfolio(mu.vec, sigma.mat, target.return)
class(e.port.msft)
e.port.msft
```

The tangency portfolio $\mathbf{t}$ is the portfolio of risky assets
with the highest Sharpe's slope and has solutions given by

$$
\mathbf{t}=\frac{\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}.
$$

To compute this portfolio with $r_{f}=0.005$ use the `tangency_portfolio()`
function:

```{r echo=TRUE}

tan.port = tangency_portfolio(mu.vec, sigma.mat, r.f)
class(tan.port)
tan.port
```

The the set of efficient portfolios of risky assets can be computed
as a convex combination of any two efficient portfolios. It is convenient
to use the global minimum variance portfolio as one portfolio and
an efficient portfolio with target expected return equal to the maximum
expected return of the assets under consideration as the other portfolio.
Call these portfolios $\mathbf{m}$ and $\mathbf{x}$, respectively.
For any number $\alpha$, another efficient portfolio can be computed
as: 
$$
\mathbf{z}=\alpha\mathbf{x}+(1-\alpha)\mathbf{m}
$$
The function `efficient_frontier()` constructs the set of efficient
portfolios using this method for a collection of $\alpha$ values
on an equally spaced grid between $\alpha_{min}$ and $\alpha_{max}$.
For example, to compute 20 efficient portfolios for values of $\alpha$
between -2 and 1.5 use:

```{r echo=TRUE}

ef = efficient_frontier(mu.vec, sigma.mat, alpha_min=-0.5, 
                        alpha_max=2, nport=20)
attributes(ef)
ef
```

Use the `summary()` method to show the weights of these portfolios.
Use the `plot()` method to create a simple plot the efficient
frontier:

```{r PortMatrixMarkowitzPlotMethod, fig.cap='Plot method for objects of class Markowitz', out.width='70%', fig.width = 6, fig.asp=.618, fig.align='center', echo=TRUE}

plot(ef)
```


The resulting plot is shown in the figure above. To create a more elaborate plot of the efficient frontier showing
the original assets and the tangency portfolio use:

```{r PortMatrixEfficientFrontierPlot, fig.cap='Efficient frontier plot.', out.width='70%', fig.width = 6, fig.asp=.618, fig.align='center', echo=TRUE}

plot(ef, plot.assets=T, col="blue", pch=16) 
points(gmin.port$sd_port, gmin.port$er_port, col="green", pch=16, cex=1.5) 
text(gmin.port$sd_port, gmin.port$er_port, labels = "Global min", pos = 4) 
points(tan.port$sd_port, tan.port$er_port, col="red", pch=16, cex=1.5) 
text(tan.port$sd_port, tan.port$er_port, labels = "Tangency", pos = 3) 
sr.tan = (tan.port$er_port - r.f)/tan.port$sd_port 
abline(a=r.f, b=sr.tan, col="green", lwd=2)
```

